---
title: Forecasting Healthy Ride ridership with prophet
author: Conor Tompkins
date: '2019-08-03'
slug: forecasting-healthy-ride-ridership-with-prophet
categories:
  - Pittsburgh
  - R
tags:
  - Pittsburgh
  - Healthy Ride
---
This post is about predicting demand for the Healthy Ride bike system in Pittsburgh. I wanted to try out Facebook's `prophet` package and try to do some time series forecasting.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

As usual, load the required packages and set up the environment:
```{r}
library(tidyverse)
library(lubridate)
library(prophet)
library(janitor)
library(hrbrthemes)

options(scipen = 999)

theme_set(theme_bw())
```

This code loads the data and formats the date column so the `prophet` package can interface with it. I use `dir()` to find all the CSV files in  the folder and then `set_names()` and `map_df(read_csv())` to read each of the CSV files into memory.
```{r}
files <- dir("data/healthy_ride/ride_data", ".csv")

df <- str_c("data/healthy_ride/ride_data/", files) %>% 
  set_names() %>% 
  map_df(read_csv) %>% 
  clean_names() %>% 
  mutate(start_date_time = starttime,
         stop_date_time = stoptime) %>%
  select(-c(starttime, stoptime)) %>% 
  gather(date_time_type, date_time, c(start_date_time, stop_date_time)) %>% 
  select(date_time_type, date_time, everything()) %>% 
  mutate(date_time_2 = date_time) %>% 
  separate(date_time, " ", into = c("date", "time")) %>% 
  mutate(id = row_number(),
         date = mdy(date),
         year = year(date),
         quarter = quarter(date),
         month = month(date, label = TRUE),
         week = week(date),
         time = hm(time), 
         hour = hour(time),
         wday = wday(date, label = TRUE),
         is_weekday = ifelse(wday %in% c("Mon", "Tues", "Wed", "Thurs", "Fri"), "weekday", "weekend"),
         yday = yday(date),
         mday = mday(date)) %>% 
  arrange(date) %>% 
  mutate(trip_duration = (tripduration / 60) / 60) %>% 
  gather(station_id_type, station_id, c(from_station_id, to_station_id)) %>% 
  gather(station_name_type, station_name, c(from_station_name, to_station_name)) %>% 
  select(date_time_type, 
         is_weekday, 
         date, 
         year,
         quarter,
         month,
         hour,
         wday,
         yday,
         mday,
         date_time_2, 
         station_id_type, 
         station_id, 
         station_name_type,
         station_name,
         everything())
```

The data I will use contains the number of rides per day and also includes the month and year. `prophet` will identify the time series patterns ("seasonality") in the data and identify the "true" pattern

`prophet` has a plug-and-play workflow that is easy to use, but it has more stringent requirements for how the data has to be shaped. The date data has to be named `ds` and the target variable has to be named `y`. I set the floor to zero because there cannot be fewer than 0 rides in a day. `prophet` requires a cap
```{r}
df <- df %>% 
  count(date) %>% 
  rename(ds = date,
         y = n) %>% 
  mutate(floor = 0,
         cap = 3 * max(y))
```

```{r}
glimpse(df)
```

This creates the set of holidays I use in the model.
```{r}
us_holidays <- prophet::generated_holidays %>% 
  as_tibble() %>% 
  filter(country == "US")
```

This code fits a model to the dataset.
```{r}
m <- prophet(df, growth = 'logistic', holidays = us_holidays)
```

`make_future_dataframe()` creates the dataframe that `prophet` uses to make its forecast. In this case, I have it create a dataframe with 365 days of additional rows to predict onto.
```{r}
future <- make_future_dataframe(m, periods = 365, freq = "day") %>% 
  mutate(floor = 0,
         cap = unique(df$cap))
```

This code performs the forecast on the future dataset.
```{r}
forecast <- predict(m, future) %>% 
  as_tibble()
```

The output is a dataframe with the date, the predicted ridership, and the upper and lower bounds of the prediction.
```{r}
forecast
```

`plot` automatically plots the `forecast` data:
```{r}
plot(m, forecast)
```

`prophet` also decomposes the various seasonal effects.
```{r}
prophet_plot_components(m, forecast) + 
  theme_bw()
```

We can of course use `ggplot` to manually plot the data.
```{r fig.height=6, fig.width=12}
df_aug <- forecast %>% 
  mutate(ds = ymd(ds)) %>% 
  left_join(df)

df_aug %>% 
  ggplot(aes(x = ds)) +
    geom_ribbon(data = df_aug %>% filter(ds > last(df$ds)), aes(ymin = yhat_lower, ymax = yhat_upper), alpha = .2, fill = "blue") +
    geom_line(data = df_aug %>% filter(ds > last(df$ds)), aes(y = yhat), color = "blue") +
    geom_point(aes(y = y), alpha = .5) +
    geom_hline(aes(yintercept = unique(df_aug$floor)), linetype = 2) +
    labs(x = NULL,
         y = "Number of rides") +
    scale_y_comma() +
    theme_ipsum()
```

`prophet` also provides functions for cross-validation.
```{r}
df_cv <- cross_validation(m, horizon = 30, units = 'days')

performance_metrics(df_cv) %>% 
  as_tibble() %>% 
  gather(metric, measure, -horizon) %>% 
  ggplot(aes(horizon, measure)) +
  geom_line() +
  facet_wrap(~metric, scales = "free_y",
             ncol = 1) +
  labs(x = "Horizon",
       y = "Measure") +
  theme_bw()
```

We can also inspect the impact of holidays on the prediction.
```{r}
df_holiday_impact <- forecast %>% 
  clean_names() %>% 
  select(ds, christmas_day:washingtons_birthday) %>% 
  select(-contains("upper"), -contains("lower")) %>% 
  pivot_longer(-ds, names_to = "holiday", values_to = "value") %>% 
  mutate(year = as.factor(year(ds))) %>% 
  filter(holiday != "holidays",
         value != 0)

df_holiday_impact %>% 
  arrange(ds) %>% 
  mutate(holiday = as.factor(holiday)) %>% 
  ggplot(aes(holiday, value, color = year)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_jitter() +
  coord_flip() +
  labs(x = "Holiday",
       y = "Impact") +
  scale_color_discrete("Year") +
  theme_ipsum()
```

## Documentation and references

* https://facebook.github.io/prophet/
* https://data.wprdc.org/dataset/healthyride-trip-data